---
title: "Genomics of CRC"
format: html
editor: visual
---

# Understanding the genomics and histological features of colorectal cancer using spatial transcriptomics

Spatial Transcriptomics allows us to identify what genes are being expressed in particular space within a tissue, allowing us to spatially analyze the types of interactions between cellular sub-populations. In this analysis, we're trying to verify how raw genetic data are processed using R, and how specific plots are generated using bioinformatic analysis.

```{r}
library(Seurat)
library(SeuratData)
library(ggplot2)
library(patchwork)
library(dplyr)

```

# Data Description

The data used is extracted from 10X Genomic's Visium HD technology. This [dataset](#0) is associated with the paper titled 'High Definition spatial transcriptomic profiling of immune cell populations in colorectal cancer.'

```{r}
Colon <- readRDS('/Users/chirag/Documents/Documents/Colorectal Cancer Research /Patient Data/colon_pca_run.rds')
Colon
```

# Data pre-processing

First, we identified the number of spots present in the raw data. Here, every spots (8X8 um hexagonal shaped spot) represent a cell in that specific cell population. We convert the raw output data into **UMI (Unique molecular identifier) count matrix.**

```{r}
plot1 <- VlnPlot(
  Colon,
  features = "nCount_Spatial",
  pt.size = 0
) +
  geom_violin(width = 0.5) +   # narrower violins → more space
  scale_x_discrete(expand = expansion(add = 0.8)) +  # increase spacing
  labs(
    y = "Spatial UMI Counts",
    x = "Cluster"
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 14, face = "bold"),
    plot.title = element_text(face = "bold", size = 15)
  ) +
  NoLegend()
plot2 <- SpatialFeaturePlot(
  Colon,
  features = "nCount_Spatial",
  alpha = c(0.1, 1)
) +
  scale_fill_viridis_c(option = "magma") +
  labs(fill = "UMI Count") +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 13, face = "bold"),
    legend.text = element_text(size = 11)
  )
final_plot <- plot1 + plot2 +
  plot_annotation(tag_levels = "A")

final_plot

```

# Clustering

We then cluster the tissue into different cell-populations. Here, each 12 spatial clusters represent specific type of cells, and the genes they are expressing. This allows us to map the interaction between different cell types.

```{r}
library(Polychrome)

n.clusters <- length(levels(Colon$Spatial_snn_res.0.4))
cluster.cols <- setNames(Polychrome::kelly.colors(n.clusters), 0:12)

p3_pub <- SpatialDimPlot(
  Colon,
  group.by = "Spatial_snn_res.0.4",
  label = FALSE,
  pt.size.factor = 6,
  image.alpha = 0, alpha = 1
) +
  scale_fill_manual(values = cluster.cols) +
  guides(fill = guide_legend(
    override.aes = list(size = 4),
    title = "Spatial clusters"
  )) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )

p3_pub

```

The following FeaturePlot show the spatial locations of **goblet cells**. The expression of the genes CLCA1, FCGBP, and MUC2 were identified in the tissue.

```{r}
genes <- c("CLCA1", "FCGBP", "MUC2")

p_feature_pub <- SpatialFeaturePlot(
  Colon,
  features = genes,
  pt.size.factor = 6,
  image.alpha = 0,
  alpha = c(0.2, 1),
  min.cutoff = "q10",
  max.cutoff = "q90",
  keep.scale = "all"
) &
  scale_fill_viridis_c(
    option = "inferno",
    name = "Normalized\nExpression"
  ) &
  theme_void() &
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )

p_feature_pub <- p_feature_pub +
  plot_annotation(tag_levels = "A")

p_feature_pub
```

Similarly, the following plot shows the expression of the genes REG1A, REG1B, CEACAM6, and TGFBI. The expressions represents the presence of **tumor cells** in the regions

```{r}
genes <- c("REG1A", "REG1B", "CEACAM6", "TGFBI")

p_feature_pub <- SpatialFeaturePlot(
  Colon,
  features = genes,
  pt.size.factor = 6,
  image.alpha = 0,
  alpha = c(0.2, 1),
  min.cutoff = "q10",
  max.cutoff = "q90",
  keep.scale = "all"
) &
  scale_fill_viridis_c(
    option = "inferno",
    name = "Normalized\nExpression"
  ) &
  theme_void() &
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )

p_feature_pub <- p_feature_pub +
  plot_annotation(tag_levels = "A")

p_feature_pub
```

Now, the following plot represents the presence of **Fibroblasts**. The expression of the genes COL1A1 and MMP2 are shown in the plots

```{r}
genes <- c("COL1A1", "MMP2")

p_feature_pub <- SpatialFeaturePlot(
  Colon,
  features = genes,
  pt.size.factor = 6,
  image.alpha = 0,
  alpha = c(0.2, 1),
  min.cutoff = "q10",
  max.cutoff = "q90",
  keep.scale = "all"
) &
  scale_fill_viridis_c(
    option = "inferno",
    name = "Normalized\nExpression"
  ) &
  theme_void() &
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )

p_feature_pub <- p_feature_pub +
  plot_annotation(tag_levels = "A")

p_feature_pub
```

Similar plots can be drawn for the cell cluster Macrophage. We can identify the expression of the genes 'LYZ1' and 'SPP1'.

```{r}
genes <- c("LYZ1", "SPP1")

p_feature_pub <- SpatialFeaturePlot(
  Colon,
  features = genes,
  pt.size.factor = 6,
  image.alpha = 0,
  alpha = c(0.2, 1),
  min.cutoff = "q10",
  max.cutoff = "q90",
  keep.scale = "all"
) &
  scale_fill_viridis_c(
    option = "inferno",
    name = "Normalized\nExpression"
  ) &
  theme_void() &
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )

p_feature_pub <- p_feature_pub +
  plot_annotation(tag_levels = "A")

p_feature_pub
```

Finally, we identify the population of Neutrophils through the expression of the gene 'SAT1'

```{r}
genes <- c("SAT1")

p_feature_pub <- SpatialFeaturePlot(
  Colon,
  features = genes,
  pt.size.factor = 6,
  image.alpha = 0,
  alpha = c(0.2, 1),
  min.cutoff = "q10",
  max.cutoff = "q90",
  keep.scale = "all"
) &
  scale_fill_viridis_c(
    option = "inferno",
    name = "Normalized\nExpression"
  ) &
  theme_void() &
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )

p_feature_pub <- p_feature_pub +
  plot_annotation(tag_levels = "A")

p_feature_pub
```

To place similar cells together in a low-dimensional space, we not visualize a UMAP plot. This method aim to preserve local distances in the dataset.

```{r}
p1_pub <- DimPlot(
  Colon,
  reduction = "umap",
  group.by = "Spatial_snn_res.0.4",
  label = TRUE,
  label.size = 4,
  pt.size = 1.2
) +
  scale_color_manual(values = cluster.cols) +
  guides(color = guide_legend(
    override.aes = list(size = 4),
    title = "Spatial clusters"
  )) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )
p2_pub <- SpatialDimPlot(
  Colon,
  group.by = "Spatial_snn_res.0.4",
  label = TRUE,
  label.size = 3,
  pt.size.factor = 6,
  image.alpha = 0
) +
  scale_fill_manual(values = cluster.cols) +
  guides(fill = guide_legend(
    override.aes = list(size = 4),
    title = "Spatial clusters"
  )) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )
final_umap_spatial <- p1_pub + p2_pub +
  plot_annotation(tag_levels = "A")

final_umap_spatial
```

```{r}
Colon@meta.data <- Colon@meta.data %>% mutate(annotations = case_when(
Spatial_snn_res.0.4 %in% 4 ~ 'Goblet', 
Spatial_snn_res.0.4 %in% 1 ~ 'Fibroblasts',
Spatial_snn_res.0.4 %in% c(5, 6, 7) ~ 'Immune Cells',
TRUE ~ as.character(Spatial_snn_res.0.4)
))
```

```{}
```

```{r}
Markers <- FindAllMarkers(Colon, group.by = "Spatial_snn_res.0.4")
head(Markers)
```
